---
title: "Clinical epi_Assignment 1"
author: "Nigus Yimer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages 

```{r}
pacman::p_load("tidyverse",
               "rio",
               "here",
               "skimr",
               "janitor")
```

## Import dataset

```{r}
GRACE <- import("GRACE_data.csv")
```

## Load packages for prediction model

```{r}
library(pROC)
library(epiDisplay)
library(PredictABEL)
```

## Explore the data

```{r}
## explore the data
names(GRACE)
head(GRACE)
str(GRACE)
```

## Describe the variables 

```{r}
## check distribution of continuous variables using plot
hist(GRACE$Oral_temp)
hist(GRACE$breaths_min)
hist(GRACE$CRP)
shapiro.test(GRACE$breaths_min)

## describe patients' characteristics--continuous variables
skim(GRACE$Oral_temp)
skim(GRACE$breaths_min)
skim(GRACE$CRP)

## describe patients' characteristics--categorical variables 
library(table1)
table1(~as.factor(pneumonia) + as.factor(Suspected_pneumonia) + as.factor(Cough_sev) + as.factor(Phlegm_colour), data= GRACE)

## tabulate by pneumonia status of patients
table1(~ as.factor(Suspected_pneumonia) + as.factor(Cough_sev) + as.factor(Phlegm_colour) | as.factor(pneumonia), data= GRACE)

```

## Change the variable types

```{r}
GRACE$Cough_sev <- as.factor(GRACE$Cough_sev)
GRACE$Phlegm_colour <- as.factor(GRACE$Phlegm_colour)
GRACE$Suspected_pneumonia <- as.factor(GRACE$Suspected_pneumonia)
```

## Univariate analysis

```{r}
null_model <- glm(pneumonia ~ 1, family = binomial, data = GRACE)
summary(null_model)

cough_uni <- glm(pneumonia ~ Cough_sev, family = binomial, data = GRACE)
summary(cough_uni)
confint(cough_uni)

temp_uni <- glm(pneumonia ~ Oral_temp, family = binomial, data = GRACE)
summary(temp_uni)

phlegm_uni <- glm(pneumonia ~ Phlegm_colour, family = binomial, data = GRACE)
summary(phlegm_uni)

breath_uni <- glm(pneumonia ~ breaths_min, family = binomial, data = GRACE)
summary(breath_uni)

sus_pnu_uni <- glm(pneumonia ~ Suspected_pneumonia, family = binomial, data = GRACE)

## check significance test using LR test

lrtest(null_model, cough_uni)
lrtest(null_model, temp_uni)
lrtest(null_model, phlegm_uni)
lrtest(null_model, breath_uni)
```

## Multivariable analysis

```{r}
full_model <- glm(pneumonia ~ Oral_temp + breaths_min, family = binomial, data = GRACE)
summary(full_model)
confint(full_model)

## convert coeffiients to odds ratios
exp(coef(full_model))
exp(confint(full_model))

```

## Model reduction using LRT

```{r}
lrtest(full_model, breath_uni)
lrtest(full_model, temp_uni)

## Build model using temperature only
full_model_1 <- glm(pneumonia ~ Oral_temp, family = binomial, data = GRACE)  # reduction is allowed at p>= 0.1

## convert estimates and CIs of categorical variables to odds ratios
exp(coef(Allvar_model))
exp(confint(Allvar_model))
```

## sign and symptoms model performance

```{r}
## Predictive probability 
prediction <- predict(full_model_1, type = "response")

## Caliberation
plotCalibration(GRACE, 7, prediction, groups = 20)

## Descrimination 
plot.roc(GRACE$pneumonia, prediction, groups= 20, main= "", ci= TRUE, print.auc= TRUE, grid.col= "red")
```

# Assess the added value of CRP in addition to signs and symptoms
```{r}
## Run the model with CRP
final_model_CRP <- glm(pneumonia ~ Oral_temp + CRP, family = binomial, data = GRACE)
## 
## Significance of additional variable
lrtest(final_model_CRP, final_model)

## Model performance
prediction_CRP <- predict(final_model_CRP, type = "response")

## Caliberation
plotCalibration(GRACE, 7, prediction_CRP, groups = 20)

## Descrimination
plot.roc(GRACE$pneumonia, prediction_CRP, groups= 20, main= "", ci= TRUE, print.auc= TRUE)
```

## Compare the two ROC curves (final model vs final model with CRP)
```{r}
## Check using plots
plot.roc(GRACE$pneumonia, prediction, main= "ROC comparison", col= "red") +
lines.roc(GRACE$pneumonia, prediction_CRP, col= "blue")

# add legend to the plot
legend("bottom", legend = c("sign and symptoms", "sign and symptoms plus CRP"), col = c("red", "blue"),lwd = 2) # add legend

## Add AUC values in the legend
auc_valuesSSCRP <- c(0.581, 0.740)
legend("bottom", legend = c(
    paste("Sign and symptoms (AUC =", auc_valuesSSCRP[1], ")"),
  paste("Sign and symptoms plus CRP (AUC =", auc_valuesSSCRP[2], ")")
 ), col = c("red", "blue"), lwd = 2)

## Check using formal tests
roc_symptom <- roc(GRACE$pneumonia, prediction, plot= TRUE, print.auc= TRUE)
roc_symptom_CRP <- roc(GRACE$pneumonia, prediction_CRP, plot= TRUE, print.auc= TRUE)
roc.test(roc_symptom, roc_symptom_CRP) 
roc.test(roc_symptom, roc_symptom_CRP, method= "bootstrap", boot.n=2000)
```

## Net reclassification improvement

```{r}
reclassification(GRACE, 7, prediction, prediction_CRP, c(0, 0.025, 0.20, 1))   

```

# Develop a prediction model based on GPs’ clinical judgement
```{r}
## Run model with suspected pneumonia
final_model_pneu <- glm(pneumonia ~ Oral_temp + Suspected_pneumonia, family = binomial, data = GRACE)

## Significance of GP's clinical judgement
lrtest(final_model_pneu, null_model)

## Predicted probability
prediction_GP <- predict(final_model_pneu, type = "response")

## Calibration
plotCalibration(GRACE, 7, prediction_GP, groups = 20)

## Descrimination
plot.roc(GRACE$pneumonia, prediction_GP, groups= 20, main= "", ci= TRUE, print.auc= TRUE)
```

# Benefit of adding CRP to the model based on GPs’ clinical judgement
```{r}
## Run model by adding CRP
final_model_pneuCRP <- glm(pneumonia ~ Oral_temp + Suspected_pneumonia + CRP, family = binomial, data = GRACE)

## Assess significance using LRT
lrtest(final_model_pneuCRP, final_model_pneu)

## Predicted probability
prediction_GP_CRP <- predict(final_model_pneuCRP, type = "response")

## Calibration
plotCalibration(GRACE, 7, prediction_GP_CRP, groups = 20)

## ## Descrimination
plot.roc(GRACE$pneumonia, prediction_GP_CRP, groups= 20, main= "", ci= TRUE, print.auc= TRUE)

```

## Compare the ROC curves
```{r}
## Check using plots
plot.roc(GRACE$pneumonia, prediction_GP, main= "ROC comparison", col= "blue") +
lines.roc(GRACE$pneumonia, prediction_GP_CRP, col= "green")
## Add legend to the plot
legend("bottom", legend = c("GPs clinical judgment", "GPs clinical judgment plus CRP"), col = c("blue", "green"),lwd = 2)
## Add AUC values in the legend
auc_valuesGPCRP <- c(0.734, 0.799)
legend("bottom", legend = c(
    paste("GPs clinical judgment (AUC =", auc_valuesGPCRP[1], ")"),
  paste("GPs clinical judgment plus CRP (AUC =", auc_valuesGPCRP[2], ")")
 ), col = c("blue", "green"), lwd = 2)

## Check using formal tests
roc_GP <- roc(GRACE$pneumonia, prediction_GP, plot= TRUE, print.auc= TRUE)
roc_GP_CRP <- roc(GRACE$pneumonia, prediction_GP_CRP, plot= TRUE, print.auc= TRUE)
roc.test(roc_GP, roc_GP_CRP) # delong method
roc.test(roc_GP, roc_GP_CRP, method= "bootstrap", boot.n=2000)  # using bootsrap method
```

## Net reclassification improvement 
```{r}
reclassification(GRACE, 7, prediction_GP, prediction_GP_CRP, c(0, 0.025, 0.20, 1))
```

